Covariate analysis of MAYO reprocessed counts (with CQN normalisation and covariates adjusted)

### 2017-04-15 code/Rmd/MAYO_REPROCESSED.Rmd
Modified Mayo reprocessing to include new samples for April freeze

knit2synapse::knitToFolderEntity(file = "./MAYO_REPROCESSED.Rmd",
                                 parentId = "syn7554308",
                                 entityName = 'MAYO Reprocessed RNASeq (Normalised)')
```

@@

```{r synapse.parameters, include=FALSE}
# Synapse parameters
parentId = 'syn7554308'
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of MAYO reprocessed RNASeq data';

@@

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_reprocessing')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data}
# Get reprocessed sample counts for temporal cortex
COUNT_TC_ID <- 'syn7987765';
ALL_USED_IDs <- COUNT_TC_ID
COUNT_TC <- read.table(synGet(COUNT_TC_ID)@filePath, header=T, sep='\t', check.names = F)
colnames(COUNT_TC) = paste(colnames(COUNT_TC), 'TCX', sep = '_')

# Get reprocessed sample counts for cerebellum
COUNT_CE_ID <- 'syn7513807';
ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_CE_ID
COUNT_CE <- read.table(synGet(COUNT_CE_ID)@filePath, header=T, sep='\t', check.names = F)

@@

# Get picard metrics from synapse
METADATA.PICARD_CE_ID <- 'syn8380392'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_CE_ID
METADATA.PICARD_CE_FPATHS <- untar(synGet(METADATA.PICARD_CE_ID)@filePath, list = FALSE, exdir = getwd())
METADATA.PICARD_CE_FPATHS <- untar(synGet(METADATA.PICARD_CE_ID)@filePath, list = TRUE, exdir = getwd())
METADATA.PICARD_CE_FPATHS = METADATA.PICARD_CE_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_CE_FPATHS)]
METADATA.PICARD_CE <- plyr::ldply(METADATA.PICARD_CE_FPATHS, read.csv)

METADATA.PICARD_CE$ID = gsub('picard/', '', METADATA.PICARD_CE_FPATHS) %>%
  gsub('/picard.CombinedMetrics.csv','',.) %>%
  sapply(function(x){str_split(x, '\\.')[[1]][1]})
colnames(METADATA.PICARD_CE) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD_CE))
colnames(METADATA.PICARD_CE) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD_CE))

METADATA.PICARD_TC_ID <- 'syn8383233'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_TC_ID
METADATA.PICARD_TC_FPATHS <- untar(synGet(METADATA.PICARD_TC_ID)@filePath, list = FALSE, exdir = getwd())
METADATA.PICARD_TC_FPATHS <- untar(synGet(METADATA.PICARD_TC_ID)@filePath, list = TRUE, exdir = getwd())
METADATA.PICARD_TC_FPATHS = METADATA.PICARD_TC_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_TC_FPATHS)]
METADATA.PICARD_TC <- plyr::ldply(METADATA.PICARD_TC_FPATHS, read.csv)

METADATA.PICARD_TC$ID = gsub('picard/', '', METADATA.PICARD_TC_FPATHS) %>%
  gsub('/picard.CombinedMetrics.csv','',.) %>%
  sapply(function(x){str_split(x, '\\.')[[1]][1]}) %>%
  gsub('Aligned','',.)
colnames(METADATA.PICARD_TC) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD_TC))
colnames(METADATA.PICARD_TC) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD_TC))

@@

### Data preprocessing
```{r preprocess.data, echo=TRUE}
# Fix metadata
METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
METADATA$Diagnosis = gsub("Pathologic Aging","Pathologic_Aging", METADATA$Diagnosis)
METADATA$BrainRegion.Diagnosis = paste(METADATA$BrainRegion, METADATA$Diagnosis, sep = '.')
METADATA$RIN2 = METADATA$RIN^2
METADATA <- METADATA %>%
  filter(!is.na(Source)) %>%
  dplyr::filter(!(ID %in% c("11311_CER", "1923_CER", "1923_TCX",
                            "11396_TCX", "11294_TCX", "11408_TCX",
                            SAMPLES.EXCLUDE)))
```
Following samples were considered outliers and removed based on PCA and/or eucledian distance: 11311\_CER, 1923\_CER, 1923\_TCX, 11396\_TCX, 11294\_TCX, 11408\_TCX
```{r preprocess.data1}
# Match covariates to expression data
indToRetain = intersect(METADATA$ID, colnames(COUNT))
indRemoved = setdiff(colnames(COUNT), METADATA$ID)

@@

### Predict missing PMI
```{r predict.pmi}
PMI = METADATA[,c('PMI', 'ID'), drop = F]
PMI$PMI = PMI$PMI * 60
rownames(PMI) = METADATA$ID

@@

# Conditional quantile normalisation
tmp.count = COUNT
genesToAnalyze = unique(intersect(rownames(GENE.GC)[!is.na(GENE.GC$percentage_gc_content)], 
                                  rownames(GENE.LEN)[!is.na(GENE.LEN$gene.length)]))
tmp.count = tmp.count[rownames(tmp.count) %in% genesToAnalyze,]
CQN.GENE_EXPRESSION = cqn(tmp.count,
                          x = GENE.GC[rownames(tmp.count), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(tmp.count), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)

# Initial normalisation of gene expression
ind = rownames(CQN.GENE_EXPRESSION$y) %in% CONTROL.PROBES$ensembl_gene_id 
EXPR = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
EXPR = EXPR[ind,]
PMI = PMI[colnames(EXPR),,drop=F]

expr = dlply(PMI, .(ID), function(mtd, expr){
  if (dim(mtd)[1] > 1)
    expr = rowMeans(expr[, mtd$ID]) %>% rownameToFirstColumn('GeneID')
  else
    expr = expr[, mtd$ID] %>% rownameToFirstColumn('GeneID')
  expr = expr %>% plyr::rename(c('DF' = unique(mtd$ID)))
  return(expr)
}, EXPR) %>%
  join_all
rownames(expr) = expr$GeneID

PMI = PMI[,colnames(PMI) %in% c('ID','PMI')] %>% unique
rownames(PMI) = PMI$ID

PMI = PMI[intersect(rownames(PMI), colnames(expr)), ]
expr = expr[,intersect(rownames(PMI), colnames(expr))]

# Split data into trainining and prediction set
x.train = expr[,!is.na(PMI$PMI)]
x.predict = expr[,is.na(PMI$PMI)]

y.train = PMI[!is.na(PMI$PMI),]

plr.model = cv.glmnet(t(x.train), y.train$PMI, family = 'gaussian')
coeff = as.matrix(coef.glmnet(plr.model, s = 'lambda.min'))
mse = sum((predict(plr.model, t(x.train), s = 'lambda.min', type = 'response') - y.train$PMI)^2)/length(y.train$PMI)
y.predict = predict(plr.model, t(x.predict), s = 'lambda.min', type = 'response') %>%
  rownameToFirstColumn('ID') %>%
  plyr::rename(c('1' = 'PMI'))

PMI = rbindlist(list(y.train, y.predict), use.names = T, fill = T)
PMI$PMI[PMI$PMI < 0] = 0

METADATA = METADATA %>% 
  dplyr::select(-PMI) %>%
  left_join(PMI)
```

@@

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=6, fig.height=6}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left')
```

@@

multiplot(plotlist = p, cols = 3)
```
### CPM Normalisation
Preprocess counts matrix using CQN (Conditional Quantile Normlaisation accounting for both GC content and gene length)

Remove genes that have less than 1 cpm counts in at least 50% of samples per BrainRegion x Diagnosis. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('SampleID') %>%
  dlply(.(BrainRegion.Diagnosis), .fun = function(mtd, count){
    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

@@

### Normalisation (with NULL)
Initial normalisation usign cqn and dispersion control with voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)

@@

# Re-compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data1.1, fig.height=6, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, scale.=T, center = T)

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('Region','Diagnosis'), sep = '\\.') %>% 
  dplyr::mutate(Region = factor(Region),
                Source = factor(Source))
plotdata$SampleID[!(plotdata$SampleID %in% c("11311_CER", "1923_CER", "1923_TCX", "11396_TCX", "11294_TCX", "11408_TCX"))] = ''

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=Source, color=Diagnosis, size=RIN))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Region~., scales = 'free_y')
p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based classification of samples
```{r decompse.normalise.data1.2, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = (COVARIATES[,c(FactorCovariates), drop = F])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

Interpretation of BrainRegion.Diagnosis
1. TCX.PSP: Progressive Supranuclear Palsy samples from Temporal cortex region
2. TCX.AD: Alzheimers Disease samples from Temporal cortex region
3. TCX.Control: Control samples from Temporal cortex region
4. TCX.Pathologic_Aging: Pathological Aging samples from Temporal cortex region
5. CER.PSP: Progressive Supranuclear Palsy samples from Cerebellum region
6. CER.AD: Alzheimers Disease samples from Cerebellum region
7. CER.Control: Control samples from Cerebellum region
8. CER.Pathologic_Aging: Pathological Aging samples from Cerebellum region

@@

# Fit contrast
contrast = makeContrasts(contrasts=c(
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.Pathologic_Aging",
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.PSP",
  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.PSP",
  "BrainRegion.DiagnosisCER.PSP-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.AD",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.Pathologic_Aging",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.PSP",
  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.AD",
  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.PSP",
  "BrainRegion.DiagnosisTCX.PSP-BrainRegion.DiagnosisTCX.AD"),
  levels = colnames(FIT$coefficients))
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:12, function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = T) %>%
    rownameToFirstColumn('ensembl_gene_id')
}, FIT.CONTR)
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis','',Comparison),
                Study = 'MAYO',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, c('from.state','to.state'), sep = '-') %>%
  tidyr::separate(from.state, c('Region','reference'), sep = '\\.') %>%
  tidyr::separate(to.state, c('Region','against'), sep = '\\.') %>%
  tidyr::unite(Comparison, reference, against, sep = '-') %>%
  left_join(GENE.PARAM %>%
              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
              unique())
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

>>>>

knit2synapse::knitToFolderEntity(file = "./MAYO_REPROCESSED.Rmd",
                                 parentId = "syn8672415",
                                 entityName = 'MAYO Reprocessed RNASeq (Normalised)')
```

@@

```{r synapse.parameters, include=FALSE}
# Synapse parameters
parentId = 'syn8672415'
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of MAYO reprocessed RNASeq data';

@@

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_reprocessing')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data}
# Get reprocessed sample counts for temporal cortex
COUNT_TC_ID <- 'syn8690799';#'syn7987765';
ALL_USED_IDs <- COUNT_TC_ID
COUNT_TC <- read.table(synGet(COUNT_TC_ID)@filePath, header=T, sep='\t', check.names = F, row.names = 1)

# Get reprocessed sample counts for cerebellum
COUNT_CE_ID <- 'syn8690904';#'syn7513807';
ALL_USED_IDs[length(ALL_USED_IDs)+1] <- COUNT_CE_ID
COUNT_CE <- read.table(synGet(COUNT_CE_ID)@filePath, header=T, sep='\t', check.names = F, row.names = 1)

@@

# Get picard metrics from synapse
METADATA.PICARD_CE_ID <- 'syn8698214';#'syn8380392'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_CE_ID
# METADATA.PICARD_CE_FPATHS <- untar(synGet(METADATA.PICARD_CE_ID)@filePath, list = FALSE, exdir = getwd())
# METADATA.PICARD_CE_FPATHS <- untar(synGet(METADATA.PICARD_CE_ID)@filePath, list = TRUE, exdir = getwd())
# METADATA.PICARD_CE_FPATHS = METADATA.PICARD_CE_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_CE_FPATHS)]
# METADATA.PICARD_CE <- plyr::ldply(METADATA.PICARD_CE_FPATHS, read.csv)
# 
# METADATA.PICARD_CE$ID = gsub('picard/', '', METADATA.PICARD_CE_FPATHS) %>%
#   gsub('/picard.CombinedMetrics.csv','',.) %>%
#   sapply(function(x){str_split(x, '\\.')[[1]][1]})
METADATA.PICARD_CE <- CovariateAnalysis::downloadFile(METADATA.PICARD_CE_ID) %>%
  dplyr::rename(ID = sample)
colnames(METADATA.PICARD_CE) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD_CE))
colnames(METADATA.PICARD_CE) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD_CE))

METADATA.PICARD_TC_ID <- 'syn8698211';#'syn8383233'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_TC_ID
# METADATA.PICARD_TC_FPATHS <- untar(synGet(METADATA.PICARD_TC_ID)@filePath, list = FALSE, exdir = getwd())
# METADATA.PICARD_TC_FPATHS <- untar(synGet(METADATA.PICARD_TC_ID)@filePath, list = TRUE, exdir = getwd())
# METADATA.PICARD_TC_FPATHS = METADATA.PICARD_TC_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_TC_FPATHS)]
# METADATA.PICARD_TC <- plyr::ldply(METADATA.PICARD_TC_FPATHS, read.csv)
# METADATA.PICARD_TC$ID = gsub('picard/', '', METADATA.PICARD_TC_FPATHS) %>%
#   gsub('/picard.CombinedMetrics.csv','',.) %>%
#   sapply(function(x){str_split(x, '\\.')[[1]][1]}) %>%
#   gsub('Aligned','',.)

METADATA.PICARD_TC <- CovariateAnalysis::downloadFile(METADATA.PICARD_TC_ID) %>%
  dplyr::rename(ID = sample)
colnames(METADATA.PICARD_TC) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD_TC))
colnames(METADATA.PICARD_TC) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD_TC))

@@

### Data preprocessing
```{r preprocess.data, echo=TRUE}
# Fix metadata
METADATA = METADATA %>%
  dplyr::mutate(AgeAtDeath = gsub("_or_above", "", AgeAtDeath),
                Diagnosis = factor(Diagnosis, 
                                   levels = c("AD", "Control", "Pathologic Aging", "PSP"), 
                                   labels = c("AD" = "AD", "Control" = "CONTROL", "Pathologic Aging" = "OTHER", "PSP" = "OTHER")),
                BrainRegion.Diagnosis = paste(BrainRegion, Diagnosis, sep = '.'),
                RIN2 = RIN^2) %>%
  filter(!is.na(Source)) %>%
  dplyr::filter(!(ID %in% SAMPLES.EXCLUDE))
```

```{r preprocess.data1}
# Match covariates to expression data
indToRetain = intersect(METADATA$ID, colnames(COUNT))
indRemoved = setdiff(colnames(COUNT), METADATA$ID)

@@

### Predict missing PMI
```{r predict.pmi}
# Get PMI
PMI = METADATA[,c('PMI', 'ID', 'Donor_ID'), drop = F]
PMI$PMI = PMI$PMI

# Get expression (convert counts to cpm)
expr = voom(COUNT, design = NULL)$E
ind = which(rowSums(expr>=1)/dim(expr)[2] >= 0.5)
expr = expr[ind,]

# Combine expression from different regions of brain of an individual into one
expr = plyr::dlply(PMI, .(Donor_ID), .fun = function(x, expr){
  data.frame(value = rowMeans(expr[,x$ID,drop = F], na.rm = T)) %>%
    plyr::rename(c('value' = unique(x$Donor_ID)))
}, expr, .parallel = F) %>%
  do.call(cbind,.)

# Split data into trainining and prediction set
PMI = dplyr::select(PMI, Donor_ID, PMI) %>% unique()
rownames(PMI) = PMI$Donor_ID
PMI = PMI[colnames(expr),]

# Fit a logistic regression model
x.train = expr[,!is.na(PMI$PMI)]
x.predict = expr[,is.na(PMI$PMI)]

y.train = PMI[!is.na(PMI$PMI),]
y.train = y.train[colnames(x.train), 'PMI']

plr.model = cv.glmnet(t(x.train), y.train, family = 'gaussian')
coeff = as.matrix(coef.glmnet(plr.model, s = 'lambda.min'))
ssres = sum((predict(plr.model, t(x.train), s = 'lambda.min', type = 'response') - y.train)^2)
sstot = sum((y.train-mean(y.train, na.rm = T))^2, na.rm = T)
R2.allgenes = 1-(ssres/sstot)

y.predict.all = predict(plr.model, t(x.predict), s = 'lambda.min', type = 'response') %>%
  rownameToFirstColumn('Donor_ID') %>%
  plyr::rename(c('1' = 'PMI'))

@@

# Initial normalisation of gene expression
ind = rownames(expr) %in% CONTROL.PROBES$ensembl_gene_id 
expr = expr[ind,]

# Fit a logistic regression model
x.train = expr[,!is.na(PMI$PMI)]
x.predict = expr[,is.na(PMI$PMI)]

y.train = PMI[!is.na(PMI$PMI),]
y.train = y.train[colnames(x.train), 'PMI']

plr.model = cv.glmnet(t(x.train), y.train, family = 'gaussian')
coeff = as.matrix(coef.glmnet(plr.model, s = 'lambda.min'))
ssres = sum((predict(plr.model, t(x.train), s = 'lambda.min', type = 'response') - y.train)^2)
sstot = sum((y.train-mean(y.train, na.rm = T))^2, na.rm = T)
R2.smallgenes = 1-(ssres/sstot)

y.predict = predict(plr.model, t(x.predict), s = 'lambda.min', type = 'response') %>%
  rownameToFirstColumn('Donor_ID') %>%
  plyr::rename(c('1' = 'PMI'))

PMI = PMI %>% 
  dplyr::select(Donor_ID, PMI) %>%
  dplyr::filter(!is.na(PMI)) %>%
  list(y.predict) %>%
  rbindlist(use.names = T, fill = T)
PMI$PMI[PMI$PMI < 0] = 0

METADATA = METADATA %>% 
  dplyr::select(-PMI) %>%
  left_join(PMI)
```
Fitted R2 values are, (i) with all gx features `r round(R2.allgenes, 3)`, (ii) with small set of inferred gx features `r round(R2.smallgenes,3)`

@@

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=10, fig.height=6}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left')
```

@@

multiplot(plotlist = p, cols = 3)
```
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per BrainRegion x Diagnosis. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
genesToAnalyze = COVARIATES %>%
  rownameToFirstColumn('SampleID') %>%
  dlply(.(BrainRegion.Diagnosis), .fun = function(mtd, count){
    processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

@@

### Outlier Analysis
```{r outlier.analysis, cache = FALSE, fig.width = 10}
indToRemove = c("11311_CER", "1923_CER", "1923_TCX", "11396_TCX", "11294_TCX", "11408_TCX")

# Find principal components of expression to plot
PC <- prcomp(voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E, scale.=T, center = T)

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  dplyr::mutate(label = SampleID) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')
plotdata$label[!(plotdata$SampleID %in% indToRemove)] = ''

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=Source, color=Diagnosis, size=RIN))
p <- p + theme_bw() + theme(legend.position="top") 
p <- p + facet_grid(BrainRegion~.+Diagnosis, scales = 'free')
p <- p + geom_text(aes(label= label), size=4, hjust=0)
p

# Plot abberent distribution of logcpm counts
tmp1 = voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E %>%
  rownameToFirstColumn('Gene.ID') %>%
  tidyr::gather(SampleID, logCPM, -Gene.ID) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID')) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')

p = ggplot(tmp1 %>%
             dplyr::filter(SampleID %in% indToRemove),
           aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'top') + facet_grid(BrainRegion~.+Diagnosis)
p

indToRetain = setdiff(colnames(PROCESSED_COUNTS$filteredExprMatrix$counts), indToRemove)
PROCESSED_COUNTS$filteredExprMatrix$counts = PROCESSED_COUNTS$filteredExprMatrix$counts[,indToRetain]
COVARIATES = COVARIATES[indToRetain,]

tmp = COVARIATES %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.') %>%
  group_by(BrainRegion, Diagnosis) %>%
  summarise(count = n()) %>%
  spread(Diagnosis, count)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples

Based on the expression pattern following samples were tagged as outliers: `r paste(indToRemove, collapse = ', ')`

Distribution of samples are: `r kable(tmp)`

### Conditional Quantile Normalisation (CQN)
Initial normalisation usign cqn and dispersion control with voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)

@@

# Re-compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
```

```{r explore.data, cache=FALSE, fig.height=5, fig.width=10}
# Plot abberent distribution of logcpm counts
tmp1 = (CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset) %>%
  rownameToFirstColumn('Gene.ID') %>%
  tidyr::gather(SampleID, logCPM, -Gene.ID) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID')) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')

p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'NONE') + facet_grid(BrainRegion+.~Diagnosis, scale = 'free')
p
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Decomposition of expression
Clustering of library normalised unadjusted data (with NULL design)
```{r decompse.normalise.data1.1, fig.height=6, fig.width=10, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, scale.=T, center = T)

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion','Diagnosis'), sep = '\\.') %>% 
  dplyr::mutate(BrainRegion = factor(BrainRegion),
                Source = factor(Source))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=Source, color=Diagnosis, size=RIN))
p <- p + theme_bw() + theme(legend.position="top") + facet_grid(BrainRegion+.~Diagnosis, scales = 'free_y')
p
```
Tree based classification of samples
```{r decompse.normalise.data1.2, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = (COVARIATES[,c(FactorCovariates), drop = F])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

Interpretation of BrainRegion.Diagnosis
1. CER.AD: Alzheimers Disease samples from Cerebellum region
2. CER.CONTROL: Control samples from Cerebellum region
3. CER.OTHER: Progressive Supranuclear Palsy and Pathological Aging samples from Cerebellum region
4. TCX.AD: Alzheimers Disease samples from Temporal cortex region
5. TCX.CONTROL: Control samples from Temporal cortex region
6. TCX.OTHER: Progressive Supranuclear Palsy and Pathological Aging samples from Temporal cortex region

@@

# Fit contrast
contrast = makeContrasts(contrasts=c(
  "BrainRegion.DiagnosisCER.AD-BrainRegion.DiagnosisCER.CONTROL",
  "BrainRegion.DiagnosisCER.OTHER-BrainRegion.DiagnosisCER.CONTROL",
  "BrainRegion.DiagnosisCER.OTHER-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisTCX.AD-BrainRegion.DiagnosisTCX.CONTROL",
  "BrainRegion.DiagnosisTCX.OTHER-BrainRegion.DiagnosisTCX.CONTROL",
  "BrainRegion.DiagnosisTCX.OTHER-BrainRegion.DiagnosisTCX.AD"),
  levels = colnames(FIT$coefficients))
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:6, function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = T) %>%
    rownameToFirstColumn('ensembl_gene_id')
}, FIT.CONTR)
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis','',Comparison),
                Study = 'MAYO',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, c('from.state','to.state'), sep = '-') %>%
  tidyr::separate(from.state, c('BrainRegion','reference'), sep = '\\.') %>%
  tidyr::separate(to.state, c('BrainRegion','against'), sep = '\\.') %>%
  tidyr::unite(Comparison, reference, against, sep = '-') %>%
  left_join(GENE.PARAM %>%
              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
              unique())
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::select(ensembl_gene_id, Comparison, Direction, BrainRegion) %>%
  group_by(BrainRegion, Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05_FC_1.2) 
kable(tmp)

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red'))
p = p + facet_grid(BrainRegion+.~Comparison, scales = 'fixed')
p
```
